Assumptions of Physics blueprint: counting evolutions as a foundation for entropy and system/state definition

### 2021-06-16 Overviews/Process entropy/ProcessEntropy.tex
ProcEntropy: another pass

\begin{abstract}
    This paper presents ideas and progress towards a more general reconceptualization of entropy that can be used to recover and extend thermodynamics and statistical mechanics. This would address foundational questions such as: why does entropy posses both objective qualities (i.e. it is a measurable property) and in (i.e. it is an ensemble property)? What is entropy maximization and why is so successful in many fields, not just physics? Is there a general notion of equilibrium that includes both dynamical and thermodynamical equilibria in a way that is relativistically invariant? The working hypothesis is that entropy is the logarithm of the count of the possible evolutions of a system within a specific process. That is, given the description of system at the desired granularity level, we count all the way the system can evolve in time that are compatible with that description. This switch from counting states to counting evolutions makes the concept more relativistically sound and allows to properly keep track of how the information about the system is transported back and forward in time by the process, which is essential to apply the concept in a more general setting. The objective is to formulate a precise mathematical framework that makes this intuition precise and recovers the usual entropies in specific settings.

@@

Previous work focused on showing how requiring experimental verifiability of a theory leads to topological and $\sigma$-algebraic structures within that theory. We have also seen how the ability to compare the level of granularity within the theory is connected to geometric and measure theoretic structures, giving us a way to quantify the information contained in each statement of the theory. We now want to characterize the relationship between the granularity of the descriptions at different times (i.e their information content. \textbf{The working hypothesis is that entropy captures how coarse or fine grained is the description of a particular system within a particular process. It does so by measuring the logarithm of the count of the possible evolutions that are compatible.} Understanding what are the appropriate basic definitions, how do they map to the standard one and how they are required for the very definitions of systems and states is therefore the central problem of this paper.

To give a sense of how this would work, we start with defining a \textbf{process domain} $\tdomain[P]$. This is a particular instance of our notion of a theoretical domain, and it will represent all meaningful statements, all descriptions at all times and at all scales, that apply to a particular process. Mathematically, it is a $\sigma$-complete Boolean algebra, which means it is closed under negation, countable disjunction (i.e. logical OR) and countable conjunction (i.e. logical AND). It will also come equipped with a partial order $\narrower$ such that $\stmt_1 \narrower \stmt_2$ if $\stmt_1 \OR \stmt_2 \equiv \stmt_2$ that indicates whether one statement gives a \textbf{narrower}, more specific, description than the other. For example:
\begin{itemize}
    \item \statement{The position of the object is between 0 and 1 meters} $\narrower$ \statement{The position of the object is between 0 and 1 kilometers}
    \item \statement{The fair die landed on 1} $\narrower$ \statement{The fair die landed on 1 or 2}
    \item \statement{The first bit is 0 and the second bit is 1} $\narrower$ \statement{The first bit is 0}
\end{itemize}

Additionally, we will have a way to compare any two statements and decide whether one provides a description at a finer level of granularity. Mathematically, we have an additional preorder $\finer : \tdomain \times \tdomain \to \mathbb{B}$. Saying $\stmt_1 \finer \stmt_2$ means that the description provided by $\stmt_1$ is \textbf{finer}, gives more information, is more precise, than the description provided by $\stmt_2$. We say that two descriptions are \textbf{equigranular}, noted $\stmt_1 \eqgran \stmt_2$, if they provide the same level of detail. For example:
\begin{itemize}
    \item \statement{The position of the object is between 0 and 1 meters} $\finer$ \statement{The position of the object is between 2 and 3 kilometers}
    \item \statement{The fair die landed on 1} $\finer$ \statement{The fair die landed on 3 or 4}
    \item \statement{The first bit is 0 and the second bit is 1} $\finer$ \statement{The third bit is 0}
\end{itemize}
In these cases, the first statement may not be contained or overlap with the second. We will assume fineness have all the properties it needs such that if a statement $\stmt[u] \in \tdomain[P]$ is chose, we can construct a measure $\mu_{\stmt[u]}$ that quantifies the granularity of a statement in terms of the unit. For example, if $\stmt[u]=$ \statement{the position of the ball is within 0 and 1 meters} and $\stmt=$ \statement{the position of the ball is within 3.5 and 5 meters}, then $\mu_{\stmt[u]}(\stmt) = 1.5$.

@@

With these basic definitions we can transport statements back forth in time and see how the process entropy changes. More specifically, for each instant of time $t \in T$ we can define a respective subdomain $\tdomain_t \subseteq \tdomain[P]$ that contains all the descriptions at that particular time. Given a statement $\stmt_0$ at time $t_0$, we can define the best prediction $\stmt_1$ at time $t_1 > t_0$ and find that its process entropy must be greater (i.e. $S(\stmt_1) \geq S(\stmt_0)$). We can also define the best reconstruction $\stmt_{-1}$ at time $t_{-1} < t_0$ and find that its process entropy also must be greater (i.e. $S(\stmt_{-1}) \geq S(\stmt_0)$). Intuitively, information about a system at a particular time can only get worse (or stay the same) as we transport that description forward or backward in time along the process. This relationship is symmetric with time. Similarly we can study how states of a particular system evolve in time, define determinism and reversibility in the terms of the ability within a process to predict or reconstruct states.

@@

TOADD: as the Boltzmann entropy fails to take into account interactions and correlations between molecules, the count of the microstates and the Shannon/Gibbs entropy fail to take into account system/environment correlations. Since these correlations may change in time, the same description of the same system will corresponds to different entropy in time.

@@

An excellent review of the history of thermodynamics and classical statistical mechanism, and therefore of the different concepts of entropy, has been compiled by Uffink (cite UFFINK). As for terminology, by Shannon entropy we will generally refer to the family of expressions based on the $p \log p$ formula. This will include the Shannon entropy, it's extension to the continuous, the Gibbs entropy and the Von Neumann entropy. By postulate entropy we will mean the one defined by the fundamental postulate of statistical mechanics, which is the logarithm of the count of microstates. We will not be discussing the Boltzmann entropy specifically, as this can be understood as a special case of the Gibbs entropy where all particles are identically and independently distributed.

@@

\textbf{Ontological vs subjective view of entropy}. On the nature of entropy, there are two positions that we find untenable. One position takes entropy to be an objective ``ontological'' quantity, ontological in the sense that, like charge or position, is an inherent property of the single instance of the system, such as total energy or total momentum. This does not work as the entropy is a property of the ensemble: for the Shannon entropy is a functional of the distribution $\rho$ and for the postulate entropy is the volume of a phase space region. The other position takes entropy to be the subjective information one has about the system, which also does not work. The ensemble is connected to possible fluctuations of the system which are measurable and independent of what we may or may not believe about the system. The idea here is that the notion of equilibrium is process dependent: it depends on both the system and the environment. It depends what happens at the boundary. Even isolation is a type of boundary, which is also technically impossible (e.g. we cannot isolate from gravitational interactions). Therefore entropy should be epistemological but objective: it quantifies the level of information (epistemological) that is accessible through a particular process allows (objective).

@@

\textbf{Equilibria and relativity}. Another problem related to time dependence is that, in general theromdynamic equilibria are not relativistic object. The notion of ``equilibria as nothing is changing'' is not invariant under boost. A volume of gas would be at equilibria only if at rest. For a boosted frame, something would be changing (i.e. the position). This also suggests an automatic interaction with gravitational forces: a volume of gas at rest would follow geodesics for other observers. Another failure of thermodynamic equilibria in a relativistic context is the following. Suppose a volume of gas is at rest for an observer reaches equilibrium at time $t_0$. This means there is a region of space $\Delta x$ that after $t_0$ can be considered at equilibrium. But equal time for a boosted frame is different, so different observers will observe different parts reaching equilibrium at different times.

@@

\textbf{Entropy of a single microstate}. The entropy of a single microstate is minus infinity, and not zero. A single microstate occupies zero entropy and the logarithm of zero is minus infinity. In terms of Gibbs/Shannon entropy, a single microstate would correspond to a delta function. The infinite density at that point contributes minus infinite entropy, and since the distribution is zero everywhere else, there are no other contributions. In quantum mechanics, all pure states have zero von Neumann/Shannon entropy. In a way, that is the problem that quantum mechanics fixes.

@@

\textbf{State space as template}. We should think of the state space as a template. We have a \textbf{state domain} $\tdomain_{X}$ which contain all statement fragments like \statement{the position of the ball is between 2 and 3 meters} together with a surjective map $\iota : \tdomain_{X} \times T \twoheadrightarrow \tdomain_t$ that adds \statement{at time t} for each possible time. The map $\iota$ must be a surjection that preserves the logical structure (i.e. logical operations, verifiability, narrowness, compatibility, ...) so that the basic logical relationships determined by the system itself are are valid at all times (i.e. \statement{at time t the position of the ball is between 2 and 3 meters} $\narrower$ \statement{at time t the position of the ball is between 0 and 300 meters} for every $t$).

Note, though, that $\iota( \cdot, t_0)$ in general is not an isomorphism (i.e. a bijection that preserves the logical structure) for two reasons. The first is that not all states will be available at every moment in every process. As we said before, if the process is dissipative the set of possible states will become smaller. In that case, $\iota$ will map all the inaccessible states to $\contradiction$, as they will be impossible. Therefore $\iota$ is not in general a bijection. The second reason is that the statement fragments in $\tdomain_{X}$ represent only descriptions about the system itself and nothing else. In a particular process, however, knowing the state of one system might tell us something about other system as well. The presence of correlations/coupling may mean that the precise knowledge of the degrees of freedom of the system may allow us to tell something about the environment or the internal structure (i.e. microstate). Mathematically, $\iota$ may map to statements that are narrower than the original fragment (e.g. we can have $\iota($\statement{position of A is 1 meter}$, t_0) \equiv$ \statement{at time $t_0$ the position of A is 1 meter and the position of B is -1 meter}).

@@

\textbf{A rich mathematical structure} To sum up, the definition of a system, then, must define several things.
\begin{itemize}
	\item A boundary, that tells us where the system ends and the environment begins
	\item A characterization of the boundary, that defines in what cases the system remains well defined
	\item A set of relaxation processes, that tells us how the system can be decoupled from the rest
	\item A set of possible independent descriptions (i.e. states), that can be characterized about the system given the circumstances
	\item A set of possible evolutions for the system (i.e. the processes) under the given circumstances.
\end{itemize}
There is a lot more work to be done to fully characterize this mathematical structure.

>>>>

\begin{abstract}
    This paper presents ideas and progress towards a more general reconceptualization of entropy that can be used to recover and extend thermodynamics and statistical mechanics. This would address foundational questions such as: why does entropy posses both objective qualities (i.e. it is a measurable property) and in (i.e. it is an ensemble property)? Is there a general notion of equilibrium that includes both dynamical and thermodynamical equilibria in a way that is relativistically invariant? What is entropy maximization, why is so successful in many fields and how is it linked to state definition? The working hypothesis is that entropy is the logarithm of the count of the possible evolutions of a system within a specific process. That is, given the description of system at the desired granularity level, we count all the way the system can evolve in time that are compatible with that description. This switch from counting states to counting evolutions makes the concept more relativistically sound and allows to properly keep track of how the information about the system is transported back and forward in time by the process, which is essential to apply the concept in a more general setting. Defining states requires, at least in some setting, the independence of the system, which indirectly leads to entropy maximization. The objective is to formulate a precise mathematical framework that makes this intuition precise and recovers the usual entropies in specific settings.

@@

Previous work focused on showing how requiring experimental verifiability of a theory leads to topological and $\sigma$-algebraic structures within that theory. Another effort is exploring how the ability to compare the level of granularity within the theory is connected to geometric and measure theoretic structures, giving us a way to quantify the information contained in each statement of the theory. We now want to characterize the relationship between the granularity  (i.e the information content) of the descriptions at different times. \textbf{The working hypothesis is that entropy captures how coarse or fine grained is the description of a particular system within a particular process. It does so by measuring the logarithm of the count of the possible evolutions that are compatible.} One way to look at it is that as the Boltzmann entropy fails to take into account interactions and correlations between molecules, the count of the microstates and the Shannon/Gibbs entropy fail to take into account system/environment correlations. Since these correlations may change in time, the same description of the same system will corresponds to different entropy in time, and this is what the process entropy is designed to solve. The central problem of this paper is therefore to understand what are the appropriate basic definitions, how do they map to the standard ones and how system independence, as characterized by entropy, is required for the very definitions of systems and states.

To give a sense of how this would work, we start with defining a \textbf{process domain} $\tdomain[P]$, which represents all meaningful statements, all descriptions at all times and at all scales, that apply to a particular process. Mathematically, it is a $\sigma$-complete Boolean algebra, which means it is closed under negation, countable disjunction (i.e. logical OR) and countable conjunction (i.e. logical AND). It will also come equipped with a partial order $\narrower$ such that $\stmt_1 \narrower \stmt_2$ if $\stmt_1 \OR \stmt_2 \equiv \stmt_2$ that indicates whether one statement gives a \textbf{narrower}, more specific, description than the other. For example:
\begin{itemize}
    \item \statement{The position of the object is between 0 and 1 meters} $\narrower$ \statement{The position of the object is between 0 and 1 kilometers}
    \item \statement{The fair die landed on 1} $\narrower$ \statement{The fair die landed on 1 or 2}
    \item \statement{The first bit is 0 and the second bit is 1} $\narrower$ \statement{The first bit is 0}
\end{itemize}

Additionally, we will have a way to compare any two statements and decide whether one provides a description at a finer level of granularity. Mathematically, we have an additional preorder $\finer : \tdomain \times \tdomain \to \mathbb{B}$. Saying $\stmt_1 \finer \stmt_2$ means that the description provided by $\stmt_1$ is \textbf{finer}, gives more information, is more precise, than the description provided by $\stmt_2$. We say that two descriptions are \textbf{equigranular}, noted $\stmt_1 \eqgran \stmt_2$, if they provide the same level of detail. For example:
\begin{itemize}
    \item \statement{The position of the object is between 0 and 1 meters} $\finer$ \statement{The position of the object is between 2 and 3 kilometers}
    \item \statement{The fair die landed on 1} $\finer$ \statement{The fair die landed on 3 or 4}
    \item \statement{The first bit is 0 and the second bit is 1} $\finer$ \statement{The third bit is 0}
\end{itemize}
In these cases, the first statement may not be contained or overlap with the second. We will assume fineness have all the properties it needs such that if a statement $\stmt[u] \in \tdomain[P]$ is chosen, we can construct a measure $\mu_{\stmt[u]}$ that quantifies the granularity of a statement in terms of the unit. For example, if $\stmt[u]=$ \statement{the position of the ball is within 0 and 1 meters} and $\stmt=$ \statement{the position of the ball is within 3.5 and 5 meters}, then $\mu_{\stmt[u]}(\stmt) = 1.5$.

@@

With these basic definitions we can transport statements back forth in time and see how the process entropy changes. More specifically, for each instant of time $t \in T$ we can define a respective subdomain $\tdomain_t \subseteq \tdomain[P]$ that contains all the descriptions at that particular time. Given a statement $\stmt_0$ at time $t_0$, we can define the best prediction $\stmt_1$ at time $t_1 > t_0$ and find that its process entropy must be greater (i.e. $S(\stmt_1) \geq S(\stmt_0)$). We can also define the best reconstruction $\stmt_{-1}$ at time $t_{-1} < t_0$ and find that its process entropy also must be greater (i.e. $S(\stmt_{-1}) \geq S(\stmt_0)$). Intuitively, information about a system at a particular time can only get worse (or stay the same) as we transport that description forward or backward in time along the process. This relationship is symmetric with time.

Similarly, we can define a trajectory $x(t)$ of a particular system, where $x$ is the full description of that system. A process is deterministic if knowing $x(t)$ means we can predict $x(t + \Delta t)$, meaning that all evolutions that start in the same state must also end in the same state: evolutions can merge and cannot diverge. Therefore the evolution count, and therefore the entropy, can never decrease: $S(x(t)) \leq S(x(t+\Delta))$. A process is reversible if knowing $x(t + \Delta t)$ means we can reconstruct $x(t)$, meaning that all evolutions that end in the same state must also start in the same state: evolutions can diverge and cannot merge. If a process is deterministic end reversible, then, the evolution count, and therefore the entropy, must be conserved $S(x(t)) = S(x(t+\Delta))$.

The last insight is that states are description of the system and only of the system. This happens only when the system is independent: the evolution of the system tells us nothing about the evolution of other systems. Independence means both factorization of the measure (and linearity of entropy) and it's maximization: the lack of correlations means all evolutions are possible. States can then be characterized by entropy in a way that is process independent, and lead to the geometrical structures of thermodynamics, classical, quantum and statistical mechanics.

@@

@@

An excellent review of the history of thermodynamics and classical statistical mechanism, and therefore of the different concepts of entropy, has been compiled by Uffink (cite UFFINK). As for terminology, by Shannon entropy we will generally refer to the family of expressions based on the $p \log p$ formula. This will include the Shannon entropy, it's extension to the continuous, the Gibbs entropy and the Von Neumann entropy. By postulate entropy we will mean the one defined by the fundamental postulate of statistical mechanics, which is the logarithm of the count of microstates. We will not be discussing the Boltzmann entropy specifically, as this can be understood as a special case of the Gibbs entropy where all particles are identically and independently distributed. When correlations between particle/molecules are present, the Boltzmann entropy fails to correspond to the thermodynamic entropy.

@@

\textbf{Ontological vs subjective view of entropy}. On the nature of entropy, there are two positions that we find untenable. One position takes entropy to be an objective ``ontological'' quantity, ontological in the sense that, like charge or position, is an inherent property of the single instance of the system, such as total energy or total momentum. This does not work as the entropy is a property of the ensemble: for the Shannon entropy is a functional of the distribution $\rho$ and for the postulate entropy is the volume of a phase space region. The other position takes entropy to be the subjective information one has about the system, which also does not work. The ensemble is connected to possible fluctuations of the system which are measurable and independent of what we may or may not believe about the system. The idea here is that the notion of equilibrium is process dependent: it depends on both the system and the environment. It depends what happens at the boundary. Even isolation is a type of boundary, which is also technically impossible (e.g. we cannot isolate from gravitational interactions). Therefore entropy should be epistemological but objective: it quantifies the level of granularity/information (epistemological) that is accessible through a particular process allows (objective).

@@

\textbf{Equilibria and relativity}. Another problem related to time dependence is that the notion of theromdynamic equilibria is not, in general, relativistic. The notion of ``equilibria as nothing is changing'' is not invariant under boost. A volume of gas would be at equilibria only if at rest. For a boosted frame, something would be changing (i.e. the position). This also suggests an automatic interaction with gravitational forces: a volume of gas at rest would follow geodesics for other observers. Another failure of thermodynamic equilibria in a relativistic context is the following. Suppose a volume of gas is at rest for an observer reaches equilibrium at time $t_0$. This means there is a region of space $\Delta x$ that after $t_0$ can be considered at equilibrium. But equal time for a boosted frame is different, so different observers will observe different parts reaching equilibrium at different times.

@@

\textbf{Entropy of a single microstate}. The entropy of a single microstate is minus infinity, and not zero. A single microstate occupies zero area and the logarithm of zero is minus infinity. In terms of Gibbs/Shannon entropy, a single microstate would correspond to a delta function. The infinite density at that point contributes minus infinite entropy, and since the distribution is zero everywhere else, there are no other contributions. In quantum mechanics, all pure states have zero von Neumann/Shannon entropy. In a way, that is the problem that quantum mechanics fixes.

@@

\textbf{State space as template}. We should think of the state space as a template we can use to instantiate groups of statements at each time. We have a \textbf{state domain} $\tdomain_{X}$ which contains all statement fragments like \statement{the position of the ball is between 2 and 3 meters} together with a surjective map $\iota : \tdomain_{X} \times T \twoheadrightarrow \tdomain_t$ that adds \statement{at time t} for each possible time. The map $\iota$ must be a surjection that preserves the logical structure (i.e. logical operations, verifiability, narrowness, compatibility, ...) so that the basic logical relationships determined by the system itself are valid at all times (i.e. \statement{at time t the position of the ball is between 2 and 3 meters} $\narrower$ \statement{at time t the position of the ball is between 0 and 300 meters} for every $t$).

Note, though, that $\iota( \cdot, t_0)$ in general is not an isomorphism (i.e. a bijection that preserves the logical structure) for two reasons. The first is that not all states will be available at every moment in every process. As we said before, if the process is dissipative the set of possible states will become smaller. In that case, $\iota$ will map all the inaccessible states to $\impossibility$, as they will be impossible. Therefore $\iota$ is not in general a bijection. The second reason is that the statement fragments in $\tdomain_{X}$ represent only descriptions about the system itself and nothing else. In a particular process, however, knowing the state of one system might tell us something about other system as well. The presence of correlations/coupling may mean that the precise knowledge of the degrees of freedom of the system may allow us to tell something about the environment or the internal structure (i.e. microstate). Mathematically, $\iota$ may map to statements that are narrower than the original fragment (e.g. we can have $\iota($\statement{position of A is 1 meter}$, t_0) \equiv$ \statement{at time $t_0$ the position of A is 1 meter and the position of B is -1 meter}).

@@

\textbf{A rich mathematical structure} To sum up, the definition of a system, then, must define several things.
\begin{itemize}
	\item A boundary, that tells us where the system ends and the environment begins
	\item A characterization of the boundary, that defines in what cases the system remains well defined
	\item A set of relaxation processes, that tells us how the system can be decoupled from the rest
	\item A set of possible independent descriptions (i.e. states), that can be characterized about the system given the circumstances
	\item A set of possible evolutions for the system (i.e. the processes) under the given circumstances.
\end{itemize}
There is a lot more work to be done to fully characterize this structure mathematically, but the current physical theory can serve as guidance. The point is that this structure has to serve as a general template that all physical theories must follow. We need a general definition of states and processes of which classical mechanics, quantum mechanics, thermodynamics and statistical mechanics are a specialization. The fact that the theories are, on the surface, so different is actually a good thing as it makes it easier to understand what elements are specific to the theories and what are more general.
